Think deep for reef resilience

### 2025-07-13 paper/paper_draft.qmd
Initial rework of discussion around carbonate budgets

Reefs in different clusters largely do not differ in mean modelled DHW levels experienced,
and exhibit overall similar DHW timeseries. While these reefs experience similar magnitude
DHWs based on SST, the proportion of the reef’s coral population impacted by heat stress
events depends on the depth of the reef. Analysis from Baird et al. [-@baird2018] is used to
calculate the degree of protection provided to a reef based on the median depth. Our results
suggest that within shallow reef ecosystems (depth \< 30m), depth contributes greatly to
reef resilience and the ability of the reef to maintain cover under heat stress. While
similar findings have been demonstrated in the eastern Pacific [@smith2014], depth refugia
analyses have often focussed on the potential of mesophotic reefs (\> 30m deep) as thermal
refugia [@bongaerts2010; @holstein2015; @semmler2017]. In contrast, Venegas et al.
[-@venegas2019] found no association between depth and the alleviation of heat stress across
a range of shallow reefs in the Pacific.
<!-- The depth refugia found on a reef in turn influences the degree of correlation between coral cover timeseries and DHW levels. Reefs in low and medium clusters have lower levels of cover – DHW correlation than those in high clusters, due to their initial sharp declines in coral cover combined with low depth – refugia in the face of persistently high thermal stress these reefs do not recover post 2030, leaving little ability to decline further in periods of high thermal stress and recover in periods of low thermal stress. -->

@@

Current reef restoration and intervention-investigation efforts on the Great Barrier Reef,
such as deployment of coral seeding and fragment devices target reefs with suitable
restoration sites in the 1-10m depth range [@howlett2021; @randall2021; @whitman2024]. One
of the main depth constraints is being able to visually identify the target site from the
surface for deployment and ongoing monitoring, in combination with factors such as
deployment method and tourism return value. Our results suggest that of the reefs analysed
within the 1-10m depth range (1800 reefs) 74% (1328) of reefs fall into the low and medium
clusters (across all GCMs at the GBRMPA management region scale), with many suffering large
future declines in reef health under climate change and possibly limiting the potential of
interventions to deliver recovery in these locations. Reef intervention planning is a
complex and challenging decision making process [@anthony2020], and our results suggest that
deployment site depth is an important factor to consider when assessing the long-term
resilience of reef sites.

@@

Furthermore, our results suggest that the capacity of reefs to maintain healthy coral
habitat by being above positive net carbonate budget levels is greater for reefs at deeper
depths than shallow depths. While the maintanence of positive carbonate budget levels does
not appear to be linked to reef connectivity metrics. The maintanence of positive carbonate
budgets is important for reef resilience under climate change.

@@

![Mean timeseries DHW values for reefs within each bioregion, separated into clusters
assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest
median area-relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters
(orange) are clusters with intermediate median area-relative cover between 2030 and 2040
within each bioregion. ‘High’ clusters (blue) are clusters with the highest median
area-relative cover between 2030 and 2040 within each bioregion. Solid lines represent the
median DHW level for each cluster. Displays results for ACCESS-ESM1-5
GCM.](../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/mean_dhw_bioregion_violin.png){#sfig-access-esm1-5-bioregion-dhwmean
width="`{julia"}
violin_width`"} <!-- ![Mean timeseries DHW tolerance value for reefs within each bioregion, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median area-relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median area-relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median area-relative cover between 2030 and 2040 within each bioregion. Solid lines represent the median DHW tolerance value for each cluster. Displays results for ACCESS-ESM1-5 GCM.](../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/dhw_tolerance_bioregion_violin.png){#sfig-access-esm1-5-bioregion-dhwtol width="`{julia}
violin_width\`"} --\>

>>>>

Reefs in different clusters largely do not differ in mean modelled DHW levels experienced,
and exhibit overall similar DHW timeseries. While these reefs experience similar magnitude
DHWs based on SST, the proportion of the reef’s coral population impacted by heat stress
events depends on the depth of the reef. Analysis from Baird et al. [-@baird2018] is used to
calculate the degree of protection provided to a reef based on the median depth. Our results
suggest that within shallow reef ecosystems (depth \< 30m), depth contributes greatly to
reef resilience and the ability of the reef to maintain positive carbonate budgets under
future conditions. While similar findings have been demonstrated in the eastern Pacific
[@smith2014], depth refugia analyses have often focussed on the potential of mesophotic
reefs (\> 30m deep) as thermal refugia [@bongaerts2010; @holstein2015; @semmler2017]. In
contrast, Venegas et al. [-@venegas2019] found no association between depth and the
alleviation of heat stress across a range of shallow reefs in the Pacific. Reefs at shallow
depths experience no, or few, years with coral cover exceeding levels likely required to
maintain positive carbonate budgets. Movement of reefs towards negative cabonate budgets
under increasing future thermal stress over time may contribute to phase shifts in reef
dominant $CaCO_3$ production from coral driven to coralline encrusting algae driven
production, or to bioerosion dominated systems with rapid erosion of physical reef
frameworks [@dudgeon2010; @perry2008].
<!-- The depth refugia found on a reef in turn influences the degree of correlation between coral cover timeseries and DHW levels. Reefs in low and medium clusters have lower levels of cover – DHW correlation than those in high clusters, due to their initial sharp declines in coral cover combined with low depth – refugia in the face of persistently high thermal stress these reefs do not recover post 2030, leaving little ability to decline further in periods of high thermal stress and recover in periods of low thermal stress. -->

@@

Current reef restoration and intervention-investigation efforts on the Great Barrier Reef,
such as deployment of coral seeding and fragment devices target reefs with suitable
restoration sites in the 1-10m depth range [@howlett2021; @randall2021; @whitman2024]. One
of the main depth constraints is being able to visually identify the target site from the
surface for deployment and ongoing monitoring, in combination with factors such as
deployment method and tourism return value. Our results suggest that of the reefs analysed
within the 1-10m depth range (1800 reefs) 74% (1328) of reefs fall into the low and medium
clusters (across all GCMs at the GBRMPA management region scale), with many unable to
maintain positive carbonate budgets under future conditions. Reef intervention planning is a
complex and challenging decision making process [@anthony2020], and our results suggest that
deployment site depth is an important factor to consider when assessing the long-term
resilience of reef sites.

@@

@@

![Mean timeseries DHW values for reefs within each bioregion, separated into clusters
assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest
median area-relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters
(orange) are clusters with intermediate median area-relative cover between 2030 and 2040
within each bioregion. ‘High’ clusters (blue) are clusters with the highest median
area-relative cover between 2030 and 2040 within each bioregion. Solid lines represent the
median DHW level for each cluster. Displays results for ACCESS-ESM1-5
GCM.](../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/mean_dhw_bioregion_violin.png){#sfig-access-esm1-5-bioregion-dhwmean
width="`{julia"} violin_width\`"}

<!-- ![Mean timeseries DHW tolerance value for reefs within each bioregion, separated into 
clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the 
lowest median area-relative cover between 2030 and 2040 within each bioregion. ‘Medium’ 
clusters (orange) are clusters with intermediate median area-relative cover between 2030 
and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest 
median area-relative cover between 2030 and 2040 within each bioregion. Solid lines 
represent the median DHW tolerance value for each cluster. Displays results for 
ACCESS-ESM1-5 GCM.](../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/dhw_tolerance_bioregion_violin.png){#sfig-access-esm1-5-bioregion-dhwtol width="`{julia}
violin_width\`"} -->
